Design and Implementation of Probabilistic Programming Language Anglican

### 2016-11-28 paper/paper.tex
normal font size
a few typos fixed

\documentclass[preprint]{sigplanconf}
\linespread{1.03}

@@

\usepackage{listings}
\usepackage{lmodern}
\ttfamily
\DeclareFontShape{OT1}{lmtt}{m}{it}
     {<->sub*lmtt/m/sl}{}
\lstdefinestyle{default}{language=Anglican, basicstyle=\ttfamily\small, columns=flexible, showstringspaces=false}

@@

The Anglican language is a subset of Clojure\footnote{It would be
possible to support almost full Clojure by expanding all macros
in the Anglican source code. However, in Clojure, unlike in
Scheme~\cite{SDF+10}, or Common Lisp~\cite{PC94}, the result
of macro-expansion of derived special forms is not well
specified and implementation specific.}.  Anglican queries
are defined within \texttt{defquery} as shown in the previous section,
using \texttt{if}, \texttt{when}, \texttt{cond}, \texttt{case},
\texttt{let}, \texttt{and}, \texttt{or}, \texttt{fn} forms;
other forms of Clojure may be supported in the future but are not now.
In \texttt{let} bindings and \texttt{fn} argument lists,
Clojure's pattern matching mechanism for vector data type
(called vector destructuring) is supported. Also, compound
literals for vectors, hash maps, and
sets are supported just like in Clojure.

@@

Providing an explicit syntax for differentiating between
Anglican and Clojure function calls would be cumbersome. Another
option would be to use meta-data to identify Anglican function
calls at runtime. However, this would impact performance, and
a good runtime performance is critical for probabilistic
programs. The approach taken by Anglican is to maintain a list
of unqualified names of primitive functions, as well of
namespaces in which all functions are primitive, and recognize
primitive functions by name --- if a function name is not in the
list, the function is an Anglican function. Global dynamically-bound 
variables \texttt{*primitive-procedures*} and
\texttt{*primitive-namespaces*} contain the initial lists of
names and namespaces, correspondingly. Of course, local bindings
can shade global primitive function names. For example,
\texttt{first} is a Clojure primitive function that takes the first
element from an ordered collection such as list and vector,
but inside the let block in the
following example, \texttt{first} is an Anglican function:
\begin{lstlisting}[style=default]
(let [first (fn [[x & y]] x)]
  (first '[1 2 3]))
\end{lstlisting}
The Anglican compiler takes care of the shading by rebinding
\texttt{*primitive-procedures*} in every lexical scope
(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
\texttt{shading-primitive-procedures} automates the shading.

@@

\paragraph{Memoization}
The author of a probabilistic model might want to 
randomly draw a feature value for each entity in a collection,
but to retain the same drawn
value for the same entity during a single run of the probabilistic
program. For example, a person may have brown or green eyes
with some probability, but the \textit{same} person will
always have the same eye color. This can be achieved
through the use of memoized functions. In Anglican, one
might write:
\begin{lstlisting}[style=default]
(let [eye-color (mem (fn [person]
                       (sample
                         (categorical
                           ['brown 0.5]
                           ['green 0.5]))))]
  (if (not= (eye-color 'bill) (eye-color 'john))
    (eye-color 'bill)
    (eye-color 'john)))
\end{lstlisting}
The \texttt{mem} form converts a function to a memoized variant,
which remembers past inputs and the corresponding outputs, and 
returns the remembered output when given such a past input, 
instead of calling the original function with it. As a result,
for every input, randon draws will be made only for the first time 
that the memoized function is called with the input; all subsequent
calls with the input will just reuse these draws return the same output.

@@

Typically, an inference
algorithm has its own implementations of
\texttt{checkpoint} for \texttt{sample}, \texttt{observe}, or
both, as well as invoke \texttt{exec} from an elaborated
conditional control flow. LMH (\texttt{anglican.lmh}) and SMC
(\texttt{anglican.smc}) are examples of inference algorithms
where either \texttt{observe} (SMC) or \texttt{sample} (LMH)
handler is overridden. In addition, SMC runs multiple particles
(corresponding to user-level threads) simultaneously, while LMH re-runs
programs from an intermediate continuation rather than from the
beginning.

@@

To overcome the above problems, Anglican uses a different scheme
which is almost as efficient as the scheme based on stack addresses for
reuse of previously drawn values, while producing addresses of constant size,
and allows manual computation of checkpoint addresses at runtime
when the default automatic scheme is insufficient. According to
the scheme:
\begin{itemize}
    \item A checkpoint may accept an auxiliary argument --- the
        checkpoint identifier. If specified, the identifier is the first
        argument of a checkpoint. For example \texttt{(sample
        'x1  (normal 0 1)} defines a \texttt{sample} checkpoint
        with identifier \texttt{x1}.
    \item If the identifier is omitted, a unique identifier is
        generated automatically as a fresh symbol.
    \item At runtime, the address of a checkpoint invocation has
        the form
        \texttt{[}\textit{checkpoint-identifier
        number-of-previous-occurrences}\texttt{]}, where the
        occurrences are of a checkpoint with the same checkpoint
        identifier.
    \item If a sequence of invocations of the same checkpoint is
        interrupted by a different checkpoint, the number of
        previous occurrences is rounded up to a multiple of
        an integer. For efficiency, a small power of 2 is used,
        such as $2^4 = 16$.
\end{itemize}

@@

\vspace{\baselineskip}
{\small
\begin{tabular}{l l}
    \textbf{old} & \textbf{new} \\ \hline
    {}\texttt{[C1 0]} & \texttt{[C1 0]} \\
    {}\texttt{[C2 0]} & \texttt{[C2 0]} \\
    {}\texttt{[C2 1]} & \textit{unused} \\
    {}\texttt{[C1 16]} & \texttt{[C1 16]} \\
    {}\texttt{[C1 17]} & \texttt{[C1 17]} \\
    {}\texttt{[C1 18]} & \textit{unused} \\
    {}\texttt{[C2 16]} & \texttt{[C2 16]} \\
    \textit{missing} & \texttt{[C2 17]} \textit{drawn}\\
    {}\texttt{[C3 0]} & \texttt{[C3 0]}
\end{tabular}
\vspace{\baselineskip}}

>>>>

\documentclass[preprint]{sigplanconf}
\linespread{1.00}

@@

\usepackage{listings}
\usepackage{lmodern}
\ttfamily
\DeclareFontShape{OT1}{lmtt}{m}{it}
     {<->sub*lmtt/m/sl}{}
\lstdefinestyle{default}{language=Anglican, basicstyle=\ttfamily, columns=flexible, showstringspaces=false}

@@

The Anglican language is a subset of Clojure\footnote{It would be
possible to support almost full Clojure by expanding all macros
in the Anglican source code. However, in Clojure, unlike in
Scheme~\cite{SDF+10} or Common Lisp~\cite{PC94}, the result
of macro-expansion of derived special forms is not well
specified and implementation specific.}.  Anglican queries
are defined within \texttt{defquery} as shown in the previous section,
using \texttt{if}, \texttt{when}, \texttt{cond}, \texttt{case},
\texttt{let}, \texttt{and}, \texttt{or}, \texttt{fn} forms;
other forms of Clojure may be supported in the future but are not now.
In \texttt{let} bindings and \texttt{fn} argument lists,
Clojure's pattern matching mechanism for vector data type
(called vector destructuring) is supported. Also, compound
literals for vectors, hash maps, and
sets are supported just like in Clojure.

@@

Providing an explicit syntax for differentiating between
Anglican and Clojure function calls would be cumbersome. Another
option would be to use meta-data to identify Anglican function
calls at runtime. However, this would impact performance, and
good runtime performance is critical for probabilistic
programs. The approach taken by Anglican is to maintain a list
of unqualified names of primitive functions, as well of
namespaces in which all functions are primitive, and recognize
primitive functions by name --- if a function name is not in the
list, the function is an Anglican function. Global dynamically-bound 
variables \texttt{*primitive-procedures*} and
\texttt{*primitive-namespaces*} contain the initial lists of
names and namespaces, correspondingly. Of course, local bindings
can shade global primitive function names. For example,
\texttt{first} is a Clojure primitive function that takes the first
element from an ordered collection such as list and vector,
but inside the let block in the
following example, \texttt{first} is an Anglican function:
\begin{lstlisting}[style=default]
(let [first (fn [[x & y]] x)]
  (first '[1 2 3]))
\end{lstlisting}
The Anglican compiler takes care of the shading by rebinding
\texttt{*primitive-procedures*} in every lexical scope
(\texttt{fn-cps}, \texttt{cps-of-let}). Macro
\texttt{shading-primitive-procedures} automates the shading.

@@

\paragraph{Memoization}
The author of a probabilistic model might want to 
randomly draw a feature value for each entity in a collection,
but to retain the same drawn
value for the same entity during a single run of the probabilistic
program. For example, a person may have brown or green eyes
with some probability, but the \textit{same} person will
always have the same eye color. This can be achieved
through the use of memoized functions. In Anglican, one
might write:
\begin{lstlisting}[style=default]
(let [eye-color (mem (fn [person]
                       (sample
                         (categorical
                           ['brown 0.5]
                           ['green 0.5]))))]
  (if (not= (eye-color 'bill) (eye-color 'john))
    (eye-color 'bill)
    (eye-color 'john)))
\end{lstlisting}
The \texttt{mem} form converts a function to a memoized variant,
which remembers past inputs and the corresponding outputs, and 
returns the remembered output when given such a past input, 
instead of calling the original function with it. As a result,
for every input, random draws will be made only for the first time 
that the memoized function is called with the input; all subsequent
calls with the input will just reuse these draws return the same output.

@@

Typically, an inference
algorithm has its own implementations of
\texttt{checkpoint} for \texttt{sample}, \texttt{observe}, or
both, as well as invokes \texttt{exec} from an elaborated
conditional control flow. LMH (\texttt{anglican.lmh}) and SMC
(\texttt{anglican.smc}) are examples of inference algorithms
where either \texttt{observe} (SMC) or \texttt{sample} (LMH)
handler is overridden. In addition, SMC runs multiple particles
(corresponding to user-level threads) simultaneously, while LMH re-runs
programs from an intermediate continuation rather than from the
beginning.

@@

To overcome the above problems, Anglican uses a different scheme
which is almost as efficient as the scheme based on stack addresses for
reuse of previously drawn values, while producing addresses of constant size,
and allows manual computation of checkpoint addresses at runtime
when the default automatic scheme is insufficient. According to
the scheme:
\begin{itemize}
    \item A checkpoint may accept an auxiliary argument --- the
        checkpoint identifier. If specified, the identifier is the first
        argument of a checkpoint. For example \texttt{(sample
        'x1  (normal 0 1))} defines a \texttt{sample} checkpoint
        with identifier \texttt{x1}.
    \item If the identifier is omitted, a unique identifier is
        generated automatically as a fresh symbol.
    \item At runtime, the address of a checkpoint invocation has
        the form
        \texttt{[}\textit{checkpoint-identifier
        number-of-previous-occurrences}\texttt{]}, where the
        occurrences are of a checkpoint with the same checkpoint
        identifier.
    \item If a sequence of invocations of the same checkpoint is
        interrupted by a different checkpoint, the number of
        previous occurrences is rounded up to a multiple of
        an integer. For efficiency, a small power of 2 is used,
        such as $2^4 = 16$.
\end{itemize}

@@

\vspace{\baselineskip}
{
\begin{tabular}{l l}
    \textbf{old} & \textbf{new} \\ \hline
    {}\texttt{[C1 0]} & \texttt{[C1 0]} \\
    {}\texttt{[C2 0]} & \texttt{[C2 0]} \\
    {}\texttt{[C2 1]} & \textit{unused} \\
    {}\texttt{[C1 16]} & \texttt{[C1 16]} \\
    {}\texttt{[C1 17]} & \texttt{[C1 17]} \\
    {}\texttt{[C1 18]} & \textit{unused} \\
    {}\texttt{[C2 16]} & \texttt{[C2 16]} \\
    \textit{missing} & \texttt{[C2 17]} \textit{drawn}\\
    {}\texttt{[C3 0]} & \texttt{[C3 0]}
\end{tabular}
\vspace{\baselineskip}}
